Supplementary material

### 2022-05-01 analysis/paper/paper.Rmd
Setting up Supplementary table

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for Norwegian Stone Age archaeology is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating, separate from both radiometric and typological methods which has been widely applied. However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the relative rarity of the method, has been under limited scrutiny compared to more traditional dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by combining independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement through Monte Carlo simulation. The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed, and compared to previous applications of the technique.

@@

The procedure for shoreline dating to be outlined is thus aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use. For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly. This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009, see also Figure \@ref(fig:exampledat)]. First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$. Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay:

@@

We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date for the occupation of the site is then:

@@

An example implementation of this approach is given in Figure \@ref(fig:shoredate), where $\lambda$ = `r round(expfit$estimate, 3)`. This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A). For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m from the site elevation $\alpha$. In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were all dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits. The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008. Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)). When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%). Only including dates modelled to be older than 2500 BCE with 95% probability, i.e. older than the Late Neolithic, improves this to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%). When only including dates older than 4000 BCE with 95% probability, i.e. only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%). The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).

>>>>

---
title: "Simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for Norwegian Stone Age archaeology is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of relative sea-level change, this offers a dating method termed shoreline dating separate from both radiometric and typological methods, which is widely applied. However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which, due to the relative rarity of the method, has been under limited scrutiny compared to more traditional dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by combining independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement through Monte Carlo simulation. The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

The procedure for shoreline dating to be outlined is aimed at determining the likely age of the occupation of a site based on its altitude above present day sea-level, with reference to shoreline displacement and the likely elevation of the site above the sea-level when it was in use. For simplicity, this is conceptually treated a single event and thus the possibility of multiple or continuous phases of occupation is not treated explicitly. This leads the problem to become analogous to that of the calibration of a radiocarbon date [@bronkramsey2009, see also Figure \@ref(fig:exampledat)]. First, finding the elevation of the sea-level at the time the site was in use is dependent on the present day elevation of the site $\alpha$ and the distance between site and the shoreline $D$. Based on the simulation results above, the distance from the elevation of the site to the contemporaneous shoreline is defined by the probability density function for exponential decay:

@@

We can then get rid of parameter $D$ by summarising over all possible distances between site and the shoreline. Given its elevation, the probability of the date of the occupation of the site is then:

@@

An example implementation of this approach is given in Figure \@ref(fig:exampledat), where $\lambda$ = `r round(expfit$estimate, 3)`. This is the decay ratio identified when considering all of the pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A). For the numerical implementation, $D$ is here defined as a sequence of increments of 0.001m, starting from the site elevation $\alpha$. In Figure \@ref(fig:shoredate) the outlined procedure is used to shoreline date all of the sites from where this relationship was derived, with the Late Neolithic sites also included for illustrative purposes. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were all dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits. The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008. Here, 100,000 age samples drawn from the probability density function of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)). When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date correspond to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%). Only including dates modelled to be older than 2500 BCE with 95% probability, i.e. older than the Late Neolithic, improves this to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%). When only including dates older than 4000 BCE with 95% probability, i.e. only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%). The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).

### 2022-05-01 analysis/supplementary-materials/supplementary.Rmd
Setting up Supplementary table

---
title: "Supplementary material"
author: "Isak Roalkvam"
date: "10/27/2021"
header-includes:
  - \usepackage{placeins}
output: 
  pdf_document:
    toc: true
bibliography: references.bib
---

@@

```{r}
# Read in radiocarbon dates to be plotted alongside analysis for
# each site.
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))
```

## Introduction

@@

## Sensitivity analysis

@@

## Site analysis

@@

The earliest ^14^C-date does not match the Middle Mesolithic inventory from Ragnhildrød [@mjaerum2012]. The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].

@@

## Numerical result of re-dating

>>>>

---
title: "Supplementary material"
author: "Isak Roalkvam"
date: "10/27/2021"
header-includes:
  - \usepackage{placeins}
output: 
  bookdown::pdf_document2:
    toc: true
    fig_caption: yes
    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
---

@@

```{r}
# Read in radiocarbon dates to be plotted alongside analysis for
# each site.
rcarbon <- read.csv(here('analysis/data/raw_data/radiocarbon.csv'))

# Read in site data for references in the final table
sitdat <- read.csv(here('analysis/data/raw_data/sites.csv'))

# Read in reported shoreline dates
prevdates <- read.csv((here(
  "analysis/data/raw_data/previous_shoreline_dates.csv"))) %>%
  mutate(start = start * -1,
         end = end * -1) %>%
  dplyr::rename("site_name" = "name")

# Load results of re-dating previously shoreline dated sites
load(here("analysis/data/derived_data/08data.RData"))
```

## Introduction {-}

@@

## Sensitivity analysis {-}

@@

## Site analyses {-}

@@

The earliest ^14^C-date, from an uncertain context, is argued to be older and not match the typological indicators in the inventory from Ragnhildrød [@mjaerum2012, 50, 71]. The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].

@@

## Numerical result of re-dating {-}

```{r}
# Find 95% HDR for shoreline dates and reorder the sitesby median date 
# (Taken directly from 08redate.R)

bdates3 <- rbind(bdates, bdates2)

hdrdat <- data.frame()
for(i in 1:length(unique(bdates3$site_name))){
  dat <-hdrcde::hdr(den = list("x" = bdates3[bdates3$site_name ==
                                                unique(bdates3$site_name)[i],
                                              "years"][[1]],
                               "y" = bdates3[bdates3$site_name ==
                                                unique(bdates3$site_name)[i],
                                              "probability"][[1]]), prob = 95)
  segdat <- data.frame(t(dat$hdr))
  groupdat <- data.frame(matrix(nrow = length(segdat[seq(1, nrow(segdat), 2),]),
                                ncol = 5))
  names(groupdat) <- c("site_name", "start", "end", "group", "year_median")

  groupdat$site_name <-  unique(bdates3$site_name)[i]
  groupdat$start <- segdat[seq(1, nrow(segdat), 2), "X95."]
  groupdat$end <- segdat[seq(2, nrow(segdat), 2), "X95."]
  groupdat$group <- seq(1:length(dat$hdr[c(TRUE, FALSE)]))
  groupdat$year_median <- dat$mode

  hdrdat <- rbind(hdrdat, groupdat)
}

hdrdat <- hdrdat[order(hdrdat$year_median, decreasing = TRUE),]
hdrdat$site_name <- factor(hdrdat$site_name, levels = unique(hdrdat$site_name))

prevdates <- prevdates[order(match(prevdates$site_name, hdrdat$site_name)),]
prevdates$site_name <- factor(prevdates$site_name,
                            levels = unique(prevdates$site_name))
prevdates$site_name <- as.character(prevdates$site_name)

tabdat <- left_join(prevdates, hdrdat, by = "site_name")

redat <- left_join(tabdat, sitdat, by = c("site_name" = "name")) %>% 
  arrange(site_name) %>% 
  group_by(site_name) %>% 
  mutate(start95 = plyr::round_any(min(start.y), 10),
         end95 = plyr::round_any(max(end.y), 10)) %>% 
  # Replace minus sign/not minus sign with BCE/CE
  mutate(start.x = ifelse(grepl("^[-]", start.x), 
            gsub("^[-]", "", start.x), 
            gsub("$", " CE", start.x)),
         end.x = ifelse(grepl("^[-]", end.x), 
            gsub("^[-](.*)", "\\1 BCE", end.x), 
            gsub("$", " CE", end.x)),
         end95 = ifelse(grepl("^[-]", end95), 
            gsub("^[-](.*)", "\\1 BCE", end95), 
            gsub("$", " CE", end95))) %>%
  mutate(start95 = ifelse(str_detect(end95, "BCE"),
                          ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1", start95), 
            gsub("$", "", start95)),
            ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1 BCE", start95), 
            gsub("$", "CE", start95)))) %>% 
  mutate(elev = paste0(min_elev, "\u2013", max_elev),
         repdate = paste0(start.x, "\u2013", end.x),
         hdrdate = paste0(start95, "\u2013", end95)) %>% 
  select(site_name, elev, repdate, hdrdate, reference)

redat[match(unique(redat$site_name), redat$site_name),] %>% 
   rowid_to_column("id") %>% 
   kable(caption = "Shoreline dating previously shorline dated sites. The range of the 95\\% HDR is rounded off to the nearest 10 years to reduce spurious precision.",
         col.names = c("", 
                     "Site",
                    "Elevation",
                    "Reported date",
                    "95% HDR",
                    "Reference"),
        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE)

\newpage

## References {-}

::: {#refs}
:::
